Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-12 WSSSPE3/report/WSSSPE3_review.tex
Section 4: minor changes and added questions.

The lightning talks were intended to give an opportunity for
attendees to quickly highlight an important issue or a potential solution. 
\begin{enumerate}
\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs. Stability:
Facilitating Research Training While Supporting Scientific Research}} A case
study: ``The Cooperative Computing Lab'' at the University of Notre Dame is a
small group of individuals whose main tasks are collaborating with people that
have large-scale computing problems, operating various parallel computer
systems, conducting computer science research, and developing open source
software. One of the main challenges they face is finding the balance between
flexibility/training and stability/quality. Their current solution for ensuring
the latter was to add a software engineer (the presenter) to the existing team
of faculty and students, who now also serves as ``spring'' between flexibility
and stability.

\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
Sustainability}} The concept of sustainability has become a topic of interest in
the field of computing, which is evidenced by the increase in the number of
events that focus on the topic. Nevertheless, it isn't well understood yet.
Penzenstadler argued that we often define sustainability too narrowly. Instead,
sustainability at its heart is a systemic concept and must be viewed from a
range of different dimensions including environmental, economic, individual,
social and technical. She introduced the Karlskrona Manifesto on Software
Design~\cite{Becker:2014}, which distills knowledge from a broad range of
related work on the topic of sustainability into a set of (mis-)perceptions and
principles. The manifesto does not proclaim that there is an easy,
one-size-fits-all solution around the corner, but rather points out that
sustainability is a ``wicked problem'' and is often misunderstood. Due to these
misperceptions, even though sustainability's importance is increasingly
recognized, many software systems are unsustainable. Even more alarming is that
most software systems' broader impacts on sustainability are unknown. To change
this, the Karlskrona Manifesto proposes nine principles and commitments. These
commitments are not dogmatic laws, but rather a commitment to rethink, to move
beyond the silo mentality, and analyze in more depth. As such, they do not
restrict, but rather open up a space for discussion.

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones,
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
to Improve Productivity of Legacy Tools}} Abani Patra presented an example how
the community using Titan2D, a Geoflow Simulation Software, increased the
productivity of their tools by improving both code and data layout. The main
obstacles in this change were the non-existence of a common version control
system for the source code, coupled with multiple versions of the same code
base, the fixed format of input files, that many input values used to be set as
compilation flags, and that the internal data layout was not suitable for modern
technologies (vectorization, accelerators). The approach of the Titan2D
developers included reinforcing the code structure using a multiple layers of
Python and C++, and a redesign of the data layout to be better suitable for
modern CPUs and accelerators.

\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
Thaney: \textit{Collaborative Software Development as Sustainable Software: Lessons
from Open Source}}
Abigail Cabunoc Mayes combines two properties of open source that, together,
create a suitable habitat for sustainable software. The first of the two
properties, public, does not only mean public code. It also includes
public discussions, a public process of including contributions, and an open
license. The second property, participatory, stresses the importance of
reaching out to the community and helping potential new members by providing
better documentation and learning experiences, like code review and examples of
good first bug reports. Together, Abigail concluded, these two properties not
only lead to higher quality, reusability and ease of understanding, but
eventually, to sustainability.

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}} Lorraine Hwang presented experiences with CIG: the Computational
Infrastructure for Geodynamics, a community software with a worldwide user base.
Like others, their main goals include high usability, sustainability, and
reproducibility. In order to achieve this, communication channels have been
developed, such as mailing lists, wikis, workshops, hackathons, tutorials, and
webinars. In order to contribute to the infrastructure, codes have to adhere to
specified minimum standards, but ideals are also given to encourage reaching the
aforementioned goals. This includes, e.g., the use of version control, certain
coding styles, the presence and nature of code tests and documentation and
certain user workflows.

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
Hwang mentioned various ways to cite efforts within the framework, including
science papers, code papers, user manuals, and the website. An analysis of the
resulting citations showed that $80\%$ of papers that use CIG codes mention the
code name, and about the same number includes a citation. Only about $20\%$
acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
cite codes (including non-CIG codes), and only about one eighth specify the
version used, showing that compared to other codes, CIG seems to be much better
cited. This, in part, is attributed to suggested citation forms communicated to
the community, and runtime tools that help collecting a relevant citation list
automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the HEP effort to Biology,
Astrophysics, Digital Curation, and other disciplines. It includes physicists,
digital librarians, as well as computer scientists, and aims to achieve some
commonality across disciplines. Examples are meta-data descriptions of archived
data, computational descriptions, descriptions of how data was processed, but
also includes questions such as whether computation replication can be
automated, and what the impact of access policies on the preservation
infrastructure is. One of the products of this effort is a suite of tools that
deals with this preservation, and the questions was posed whether that software
itself is sustainable. Points that were brought up included the need for a user
community depending on a given software, and the need to provide added value for
its users. An important method to achieve this was to work with the user
community from the start, and to budget that way. For the specific example of
the preservation software, this means that besides adding value to the
community, it needs to be transparent to their workflow, i.e., not requiring
additional effort to preserve ``research objects''.

\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
Susheel Varma: \textit{An update from UK Research Software Engineers}} James
Hetherington started by listing some of the problems research software faces,
which include low levels of reuse and poor standard of verification. For a long
time, technical solutions to such technical problems were focus of the eResearch
community, including research software distributions, grids, middleware and
workflows. Some limited adoption can be seen today in research communities, but
the main problems have not been solved to a sufficient level. James Hetherington
hypothesizes that instead of technical solutions, social innovation is needed: a
new role in the academic system focused on research software that combines the
best parts of a craftsperson and a scholar. However, social innovation in
centuries-old institutions is hard. Alternatives to such a new role would have
to include rewards for good research software, recognition of software as
academic output, and rejecting submissions based on irreproducible computational
results. Some advantages of research software engineering (RSE) groups include
the possibility of training in reproducible computational research, providing
collaborations for for researchers who don't want to be programmers, and
synergies with research computing platforms. The success of such a group would
be measured by the output and quality of the research software, and could be
part of Research Computing or faculty. Members would not be independent
researchers, but would have to have a research background. An attempt to form a
community of RSE groups within the UK has been underway for several years,
including funding from the UK Research Council. An open question is whether this
approach can be adopted also in other countries, including the USA.

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}} Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, great developers, good design, and software
engineering practices, however, Dan placed the users. He explained this using a
traditional software development model starting from gathering requirements, and
reaching release through a design and development and testing phase. The main
deficiency that with this approach was pointed out to be the too-late
interaction with users. Instead, an alternative approach was proposed that, at
first, skips the development phase and repeatedly goes through requirement
gathering, design and user interaction/learning phases, and only eventually
starts development once an agreement is reached, leading to an increased user
satisfaction, higher adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, and Sergei
Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
Biology: Lessons Learned on the KBase Project}} In this talk, Dan Gunter
presented experiences and lessons learned as part of KBase, an open software and
data platform for addressing the grand challenge of systems biology: predicting
and designing biological function. KBase is a unified system that integrates
data and analytical tools for comparative functional genomics of microbes,
plants, and their communities. However, it is also a collaborative environment
for sharing methods and results, and placing those results in the context of
knowledge in the field. Being a large, multi-institutional project, one of the
big challenges is to agree on standard to enable a single, maintainable system.
Working in isolation does not work (anymore) within this field, and they also
didn't have standards for software engineering. This is contrasted to computer
science research, where these approaches shorten design cycles, leading to more
time for highly rewarded activities like publishing, performance studies,
graduating students, or protecting ideas before publication. Sustainability is
not of big concern there, compared to these. \katznote{prev sentence is unclear}
Instead of this more traditional approach, KBase uses the following for large,
distributed teams. After picking projects and team members, 4-5 teams work on
projects about two weeks before a 1-week evaluation by the executive committee,
and after which new teams and projects might be chosen to restart the ``agile''
development cycle. This process is intentionally open and documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
approach to overcome some hurdles in scientific publishing: the disconnect
between experimental data, research software, and publications. A current effort
in the Geosciences is the ``Geoscience Paper of the Future,'' which includes
four elements. First, it forms a modern paper including text, data, and pointers
to supplementary material. Second, it is reproducible, including data
processing, workflow, and visualization tools. Third, it is part of open
science, which includes being publicly available under open licenses, and
providing meta-data. Last, but not least, it uses digital scholarship elements
like persistent identifiers for software, data, and authors, and it cites both
data and software. OntoSoft is an effort to help this effort, providing software
stewardship for the Geosciences. As part of this initiative, a special issue of
a journal in a Geoscience areas is planned that only includes Geoscience papers
of the future, with submissions open until the end of 2015. In addition,
training sessions are provided to geoscientists to learn best practices in
software and data sharing, provenance documentation, and scholarly publication.

\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
Accreditation Framework}} Neil Chue Hong presented a proposal to build a
scientific software accreditation framework. One of the aims of such a framework
would be to measure how ``good'' a given piece of software is, and to evaluate
how this can be effectively measured in the first place. This can be compared to
the effort of the standardized, easy to read, and understandable nutritional
labeling of goods, which only contain a small set of categories. However, such a
framework for software would be more difficult due to different existing
community norms. The challenges such a framework faces include that many
measurements are subjective, many metrics are too costly, and that
self-assessment needs to be encouraged. Possible categories would include
availability, usability, transferability, among others. Such a framework could
enable both improvement of specific software, as well as comparisons of similar
software. An accreditation by such a framework could then be part of software
management plans, ensuring that software is accessible and reusable throughout
the proposed project and beyond.

>>>>

The lightning talks were intended to give an opportunity for attendees to
quickly highlight an important issue or a potential solution.
%
\begin{enumerate}
\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
Facilitating Research Training While Supporting Scientific Research}} Benjamin Tovar presented a case
study of ``The Cooperative Computing Lab'' at the University of Notre Dame, which is a
small group of individuals whose main tasks are collaborating with people that
have large-scale computing problems, operating various parallel computer
systems, conducting computer science research, and developing open source
software. One of the main challenges they face is finding a balance between
flexibility/training and stability/quality. Their current solution for ensuring
the latter was to add a software engineer (the presenter) to the existing team
of faculty and students, who now also serves as a ``spring'' between flexibility
and stability.

\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
Sustainability}} The concept of sustainability has become a topic of interest in
the field of computing, which is evidenced by the increase in the number of
events that focus on the topic. Nevertheless, it isn't well understood yet.
Birgit Penzenstadler argued that we often define sustainability too narrowly. Instead,
sustainability at its heart is a systemic concept and must be viewed from a
range of different dimensions including environmental, economic, individual,
social, and technical. She introduced the Karlskrona Manifesto on Software
Design~\cite{Becker:2014}, which distills knowledge from a broad range of
related work on the topic of sustainability into a set of (mis-)perceptions and
principles. The manifesto does not proclaim that there is an easy,
one-size-fits-all solution around the corner, but rather points out that
sustainability is a ``wicked problem'' and is often misunderstood. Due to these
misperceptions, even though sustainability's importance is increasingly
recognized, many software systems are unsustainable. Even more alarming is that
most software systems' broader impacts on sustainability are unknown. To change
this, the Karlskrona Manifesto proposes nine principles and commitments. These
commitments are not dogmatic laws, but rather commitments to rethink, to move
beyond the silo mentality, and to analyze in more depth. As such, they do not
restrict, but rather open up a space for discussion.

\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones,
and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
to Improve Productivity of Legacy Tools}} Abani Patra presented an example of how
the community using Titan2D, a Geoflow Simulation Software, increased the
productivity of their tools by improving both code and data layout. The main
obstacles in this change were the non-existence of a common version control
system for the source code, coupled with multiple versions of the same code
base, the fixed format of input files, that many input values were set as
compilation flags, and that the internal data layout was not suitable for modern
technologies (vectorization, accelerators). The approach of the Titan2D
developers included reinforcing the code structure using  multiple layers of
Python and C++ interfaces, and a redesign of the data layout to be more suitable for
modern CPUs and accelerators.

\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
Thaney: \textit{Collaborative Software Development as Sustainable Software: Lessons
from Open Source}}
Abigail Cabunoc Mayes combines two properties of open source that, together,
create a suitable habitat for sustainable software. The first of the two
properties, public, does not only mean public code. It also includes
public discussions, a public process of including contributions, and an open
license. The second property, participatory, stresses the importance of
reaching out to the community and helping potential new members by providing
better documentation and learning experiences, like code review and examples of
good first bug reports. Together, Abigail concluded, these two properties not
only lead to higher quality, reusability, and ease of understanding, but also
eventually, to sustainability.

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}} Lorraine Hwang presented experiences with CIG: the Computational
Infrastructure for Geodynamics, a community software with a worldwide user base.
Like others, their main goals include high usability, sustainability, and
reproducibility. In order to achieve these goals, communication channels have been
developed, such as mailing lists, wikis, workshops, hackathons, tutorials, and
webinars. In order to contribute to the infrastructure, codes have to adhere to
specified minimum standards, but ideals \choinote{ideals or ideas?} are also given to encourage reaching the
aforementioned goals. These include, e.g., the use of version control, certain
coding styles, the presence and nature of code tests and documentation, and
certain user workflows.

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
Hwang mentioned various ways to cite efforts within the framework, including
science papers, code papers, user manuals, and the CIG website. An analysis of the
resulting citations showed that $80\%$ of papers that use CIG codes mention the
code name, and about the same number includes a citation. Only about $20\%$
acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
cite codes (including non-CIG codes), and only about one eighth specify the
version used. Compared to other codes, CIG seems to be much better
cited. This, in part, is attributed to suggested citation forms communicated to
the community, and runtime tools that help collecting a relevant citation list
automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the HEP \choinote{What is HEP?} effort to Biology,
Astrophysics, Digital Curation, and other disciplines. It includes physicists,
digital librarians, as well as computer scientists, and aims to achieve some
commonality across disciplines. Examples are meta-data descriptions of archived
data, computational descriptions, descriptions of how data was processed, 
questions such as whether computation replication can be
automated, and what the impact of access policies on the preservation
infrastructure is. One of the products of this effort is a suite of tools that
deals with this preservation, and the questions was posed whether that software
itself is sustainable. Points that were brought up included the need for a user
community depending on a given software, and the need to provide added value for
its users. An important method to achieve this was to work with the user
community from the start, and to budget that way. For the specific example of
the preservation software, this means that besides adding value to the
community, it needs to be transparent to their workflows, i.e., not requiring
additional effort to preserve ``research objects''.

\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
Susheel Varma: \textit{An update from UK Research Software Engineers}} James
Hetherington started by listing some of the problems research software faces,
which include low levels of reuse and poor standard of verification. For a long
time, technical solutions to such technical problems were focus of the eResearch
community, including research software distributions, grids, middleware and
workflows. Some limited adoption can be seen today in research communities, but
the main problems have not been solved to a sufficient level.  Hetherington
hypothesized that instead of technical solutions, social innovation is needed: a
new role in the academic system focused on research software that combines the
best parts of a craftsperson and a scholar. However, social innovation in
centuries-old institutions is hard. Alternatives to such a new role would have
to include rewards for good research software, recognition of software as
academic output, and rejecting submissions based on irreproducible computational
results. Some advantages of research software engineering (RSE) groups include
the possibility of training in reproducible computational research, providing
collaborations for researchers who don't want to be programmers, and
synergies with research computing platforms. The success of such a group would
be measured by the output and quality of the research software, and could be
part of Research Computing or faculty. Members would not be independent
researchers, but would have to have a research background. An attempt to form a
community of RSE groups within the UK has been underway for several years,
including funding from the UK Research Council. An open question is whether this
approach can be adopted in other countries, including the USA.

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}} Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, great developers, good design, and software
engineering practices, Dan placed the users. He explained this using a
traditional software-development model starting from gathering requirements, and
reaching release through a design and development and testing phase \choinote{very complicated phrase of the phase}. The main
deficiency with this approach was pointed out to be the too-late
interaction with users. Instead, an alternative approach was proposed that, at
first, skips the development phase and repeatedly goes through requirement
gathering, design and user interaction/learning phases, and only eventually
starts development once an agreement is reached, leading to an increased user
satisfaction, higher adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, and Sergei
Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
Biology: Lessons Learned on the KBase Project}} In this talk, Dan Gunter
presented experiences and lessons learned as part of KBase, an open software and
data platform for addressing the grand challenge of systems biology: predicting
and designing biological function. KBase is a unified system that integrates
data and analytical tools for comparative functional genomics of microbes,
plants, and their communities. However, it is also a collaborative environment
for sharing methods and results, and placing those results in the context of
knowledge in the field. Being a large, multi-institutional project, one of the
big challenges is to agree on standard to enable a single, maintainable system.
Working in isolation does not work (anymore) within this field, and they \choinote{who are they?} also
didn't have standards for software engineering. This is contrasted to computer
science research, where these approaches shorten design cycles, leading to more
time for highly rewarded activities like publishing, performance studies,
graduating students, or protecting ideas before publication. Sustainability is
not of big concern there, compared to these. \katznote{prev sentence is unclear}
Instead of this more traditional approach, KBase uses the following for large,
distributed teams. After picking projects and team members, 4-5 teams work on
projects about two weeks before a 1-week evaluation by the executive committee,
and after which new teams and projects might be chosen to restart the ``agile''
development cycle. This process is intentionally open and documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
approach to overcome some hurdles in scientific publishing: disconnects
between experimental data, research software, and publications. A current effort
in the Geosciences is the ``Geoscience Paper of the Future,'' which includes
four elements. First, it forms a modern paper including text, data, and pointers
to supplementary materials. Second, it is reproducible, including data
processing, workflow, and visualization tools. Third, it is part of open
science, which includes being publicly available under open licenses, and
providing meta-data. Last, but not least, it uses digital scholarship elements
like persistent identifiers for software, data, and authors, and it cites both
data and software. OntoSoft is a tool to help this effort, providing software
stewardship for the Geosciences. As part of this initiative, a special issue of
a journal in Geoscience areas is planned to include only Geoscience papers
of the future, with submissions open until the end of 2015. In addition,
training sessions are provided to geoscientists to learn best practices in
software and data sharing, provenance documentation, and scholarly publication.

\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
Accreditation Framework}} Neil Chue Hong presented a proposal to build a
scientific software accreditation framework. One of the aims of such a framework
would be to measure how ``good'' a given piece of software is, and to evaluate
how this can be effectively measured in the first place. This can be compared to
the effort of the standardized, easy to read, and understandable nutritional
labeling of food, which only contain a small set of categories. However, such a
framework for software would be more difficult due to different existing
community norms. The challenges such a framework faces include that many
measurements are subjective, that many metrics are too costly, and that
self-assessment needs to be encouraged. Possible categories would include
availability, usability, transferability, among others. Such a framework could
enable both improvement of specific software, as well as comparisons of similar
software. An accreditation by such a framework could then be part of software
management plans, ensuring that software is accessible and reusable throughout
the proposed project and beyond.
